---
title: "Intuitive Immune Isolation of cell types with scGate. Basic examples"
author: "M. Andreatta, A. Berenstein and S. Carmona"
date: "29/09/2021"
output:
  rmdformats::readthedown:
    self-contained: true
    highlight: haddock
    thumbnails: false
    css: styles.css
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file, encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'scGate.basic.usage.html'))})
---


```{r, message=F, warning=F,results=F}
renv::restore()
library(ggplot2)
library(dplyr)
library(patchwork)
remotes::install_github("carmonalab/scGate",ref = "dev")
library(scGate)
```


```{r setup, echo=FALSE, message=F, warning=F, results=F}
library(knitr)
library(rmdformats)
library(formatR)


## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               cache.lazy=FALSE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE,
               dev='png')
opts_knit$set(width=75)

```




## Load some testing datasets and models

Load default models from DB
```{r,collapse=TRUE}
models.DB <- scGate::get_scGateDB()

names(models.DB$human$generic)

names(models.DB$mouse$generic)

```

or alternatively, you can load models from your own personalized folder
```{r}
# models_path = '/home/user/custom_model'
# models <- scGate::load_scGate_model(models_path) 
```

### Have a look at some model (for B cells and plasma cells)
```{r message=F,results=F}
my_scGate_model <- models.DB$human$generic$PanBcell
plt.tree <- scGate::plot_tree(my_scGate_model)
plt.tree
```



## Run an example
```{r}
testing.datasets <- scGate:::get_testing_data(version = 'hsa.latest')
```

```{r,collapse =T}
# List available testing datasets:
names(testing.datasets)
# Select one
dset <- "Zilionis"
obj <- testing.datasets[[dset]]
```


```{r, collapse=T}
# run scGate
obj <- scGate(obj, model = my_scGate_model, assay = DefaultAssay(obj))
DimPlot(obj, cols = c(list(Impure = "gray", Pure = "#00ae60")))
```

## Visualize filtering results by layer

```{r,fig.height=8,fig.width=10 }
plots <- scGate::plot_levels(obj)
plots[["annotation"]] <- DimPlot(obj, group.by = "cell_type",label = T, repel =T,label.size = 3) + ggtitle("Author's manual annot") + NoLegend() 
wrap_plots(plots, ncol = 2)
```

## Evaluate performance of filtering compared to original annotation
```{r,collapse=T}
perf.scGate = scGate::performance.metrics(actual = obj$PanBcell, pred = obj$is.pure== "Pure")
perf.scGate
```

## fast evaluation of model modifications
```{r,fig.width=16,fig.height=12,collapse = T, message = F}
# first, get model performance for a current model on available testing data
my_scGate_model <- models.DB$human$generic$PanBcell
panBcell.performance <- test_my_model(my_scGate_model, target = "PanBcell")
panBcell.performance$performance
```

Now, you can perform some model modifications
```{r,collapse=T, message =F}
# Some (absurd) model modification
my.custom.model <- my_scGate_model
my.custom.model%>%tail(3)
## REMOVE negative Tcell layer in last level (it make no sense, but we do it only for demonstrative purposes)
my.custom.model <- my.custom.model[-18,]   
my.custom.model%>%tail(3)
```

And evaluate this consecuences in a fast way.
```{r,fig.width=16,fig.height=12,collapse = T, message = F}
## Evaluate how the new model performs
new.model.performance <- test_my_model(my.custom.model, testing.version = 'hsa.latest',target = "PanBcell",PLOT = F)
new.model.performance$performance  

```
Notice: dropping a relevant negative signature in the last level of your model, do not affect model recall, but precision get worst in all the provided datasets.





