---
title: "Intuitive Immune Isolation of cell types with scGate. Basic examples"
author: "M. Andreatta, A. Berenstein and S. Carmona"
date: "29/09/2021"
output:
  rmdformats::readthedown:
    self-contained: true
    highlight: haddock
    thumbnails: false
    css: styles.css
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file, encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'scGate.basic.usage.html'))})
---


```{r, message=F, warning=F,results=F}
renv::restore()
library(ggplot2)
library(dplyr)
library(patchwork)
remotes::install_github("carmonalab/scGate",ref = "dev")
library(scGate)
```


```{r setup, echo=FALSE, message=F, warning=F, results=F}
library(knitr)
library(rmdformats)
library(formatR)


## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               cache.lazy=FALSE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE,
               dev='png')
opts_knit$set(width=75)

```


## Load some testing datasets 
```{r}
testing.datasets <- scGate:::get_testing_data(version = 'hsa.latest')
```

Let's play with a PBMC dataset ("Satija") from the publication XXX
```{r,collapse =T}
obj <- testing.datasets[["Satija"]]
DimPlot(obj, label = T, repel = T, group.by = "cell_type") + theme(legend.position = "none", aspect.ratio = 1) 
```

Set up a simple scGate gating model to purify a population of interest, for instance, B cells.
B cells are usually identified by the expression of CD20, encoded by MS4A1

## Simple gating model definition
```{r}
# model initizlization
my_scGate_model <- edit_model()  

# add a first signature
my_scGate_model <- edit_model(model = my_scGate_model, name = "Bcell", signature = c("MS4A1"))
my_scGate_model
```

## Purify cell population of interest with scGate
```{r, collapse=T}
# run scGate
obj <- scGate(data = obj, model = my_scGate_model, assay = DefaultAssay(obj))
DimPlot(obj, cols = c(list(Impure = "gray", Pure = "green"))) + theme(aspect.ratio = 1) 
```
Because in this demo example we know the ground truth (at least, as defined by the authors), let's evaluate scGate predictive performance in terms of precision/positive predictive value, recall/sensitivity, and accuracy using Matthews Correlation Coefficient (MCC)
```{r}
scGate::performance.metrics(grepl("^B ",obj$cell_type),obj$is.pure=="Pure")
```
With this very simple model, >99% (Recall) are isolated with a >99% purity (Precision), and and overall accuracy/MCC >99%


# Another example for gating NKs, which express KLRD1. However, KLRD1 can also be expreseed by some T cell subsets, so we can filter these out by including "negative" T cell markers such as CD3D or CD3G
```{r}
# simple model building 
my_scGate_model <- edit_model()  # model initizlization
my_scGate_model <- edit_model(model = my_scGate_model, name = "NK", signature = c("KLRD1"))  # add one positive signature
my_scGate_model <- edit_model(model = my_scGate_model, name = "Tcell", signature = c("CD3D","CD3G"),negative = T)  # add one negative signature

# run the model
obj <- scGate(data = obj, model = my_scGate_model, assay = DefaultAssay(obj))
DimPlot(obj, cols = c(list(Impure = "gray", Pure = "green"))) + theme(aspect.ratio = 1) 
```

Let's evaluate performance
```{r}
scGate::performance.metrics(grepl("^NK",obj$cell_type),obj$is.pure=="Pure")
```

With this simple model, >85% of NKs (recall) are isolated with a 100% purity (Precision=1), and overall accuracy/MCC of 92%


Another example for gating plasmacytoid dendritic cells (pDCs), which can be isolated using marker LILRA4 (eg https://pubmed.ncbi.nlm.nih.gov/30395816/)
```{r}
# simple model building 
my_scGate_model <- edit_model()  # model initizlization
my_scGate_model <- edit_model(model = my_scGate_model, name = "pDC", signature = c("LILRA4"))  # add one positive signature

# run the model
obj <- scGate(data = obj, model = my_scGate_model, assay = DefaultAssay(obj))
DimPlot(obj, cols = c(list(Impure = "gray", Pure = "green"))) + theme(aspect.ratio = 1) 
```
```{r}
scGate::performance.metrics(obj$cell_type=="pDC",obj$is.pure=="Pure")
```


Here we have been using a dataset derived from blood (PBMC). When working with more complex tissues, such as tumors, we might need to apply more stringent criteria. That's why scGate allows to purify by steps, using a hierarchical gating model.

Let's explore the whole tumor dataset by Jerby-Arnon et al (CITE)
```{r,collapse =T}
obj <- testing.datasets[["JerbyArnon"]]
DimPlot(obj, label = T, repel = T, group.by = "cell_type") + theme(legend.position = "none", aspect.ratio = 1) 
```
Here we have non-immune populations such as malignant/cancer cells (Mal), Endothelial cells (Endo) and cancer-associated fibroblasts (CAF).
We can try to purify a broadly defined cell population, for instance, all immune cells, using pan-immune cell marker CD45 encoded by PTPRC:

```{r}
my_scGate_model <- edit_model()  # model initizlization
my_scGate_model <- edit_model(model = my_scGate_model, name = "immune", signature = c("PTPRC")) 
obj <- scGate(data = obj, model = my_scGate_model, assay = DefaultAssay(obj))
DimPlot(obj, cols = c(list(Impure = "gray", Pure = "green"))) + theme(aspect.ratio = 1) 
```

```{r}
scGate::performance.metrics(obj$cell_type %in% c("B.cell","NK","T.CD4","T.CD8","T.cell","Macrophage"),obj$is.pure=="Pure")
```

Now let's try to recover macrophages (eg using marker CD68). Instead of purifying macrophages from the whole tissue, we can do that in two steps:
first immune cells as in the previous example, then macrophages. Hierarchical gating models can be specified in scGate using parameter "level"  as follows:

```{r}
my_scGate_model <- edit_model()  # model initizlization
my_scGate_model <- edit_model(model = my_scGate_model, name = "immune", signature = c("PTPRC"), level = 1)  # add positive signature at first step
my_scGate_model <- edit_model(model = my_scGate_model, name = "macrophage", signature = c("CD68","FCGR1A"), level = 2)  # add positive signature at second step
obj <- scGate(data = obj, model = my_scGate_model, assay = DefaultAssay(obj))
DimPlot(obj, cols = c(list(Impure = "gray", Pure = "green"))) + theme(aspect.ratio = 1) 
```

```{r}
scGate::performance.metrics(obj$cell_type %in% c("Macrophage"),obj$is.pure=="Pure")
```

Here we isolated 92% of macrophages with a 90% purity (according to annotation by Jerby-Arnon et al). We could easiliy improve PREC and REC by adding more positive and negative markers, respectively (e.g. removing lymphocytes)


TODO: add + theme(aspect.ratio = 1) to plots_levels and Pure color to 'green'

We can see how the two-level gating model worked in each step using scGate's plot_levels.
We see on the left the immune cell purification step, and on the right, the macrophage purifications step.
```{r}
wrap_plots(plot_levels(obj))
```

scGate models of arbitrary complexity can be easily constructed to achieve high purification performance.
These gating models can be written using the edit_model(), and manually edited using eg R's fix().
You can also export your basic model created with edit_model() as tabulated text and edit on Excel
```{r}
fix(my_scGate_model) # edit locally in R
write.table(my_scGate_model,"test.tsv",sep="\t") # export and then edit on Excel
my_scGate_model_edited <- scGate::load_scGate_model("test.tsv") # tabulated-text model can loaded
```


We provide a built-in function `test_my_model` to automatically evaluate the performance of a gating model using 3 pre-annotated testing datasets:

## fast evaluation of model modifications
```{r,fig.width=16,fig.height=12,collapse = T, message = F}
my_scGate_model <- edit_model()  
my_scGate_model <- edit_model(model = my_scGate_model, name = "Bcell", signature = c("MS4A1"))
panBcell.performance <- test_my_model(my_scGate_model, target = "Bcell")
panBcell.performance$performance
```

We see that in one of the datsets this simple model did not have an excellent precision (85%).
We can refine it, for instance by removing potentially contaminat T cell genes (CD3D) and try again
```{r}
my_scGate_model <- edit_model(model = my_scGate_model, name = "Tcell", signature = c("CD3D"),negative = T)
panBcell.performance <- test_my_model(my_scGate_model, target = "Bcell")
panBcell.performance$performance

```
Now performance is very high for the three datasets.

We also provide some pre-defined gating models, these can be retrieved with get_scGateDB()

Load default models from DB
```{r,collapse=TRUE}
models.DB <- scGate::get_scGateDB()

names(models.DB$human$generic)

names(models.DB$mouse$generic)

```

### Have a look at some model (PanBcell for B cells and plasma cells)

To visualize complex models in tree-like structures, we provide plot_tree() function
This is a model to purify B cells and plasma cells with high accuracy

```{r message=F,results=F}
my_scGate_model <- models.DB$human$generic$Plasma_cell
plt.tree <- scGate::plot_tree(my_scGate_model)
plt.tree
```


Run an example, now on another tumor dataset by Zilionis et al.

```{r}
obj <- testing.datasets[["Zilionis"]]
DimPlot(obj, label = T, repel = T, group.by = "cell_type") + theme(legend.position = "none", aspect.ratio = 1) 
```


```{r, collapse=T}
# run scGate
obj <- scGate(obj, model = my_scGate_model)
DimPlot(obj, cols = c(list(Impure = "gray", Pure = "green"))) + theme(aspect.ratio = 1) 
```

## Visualize filtering results by layer

```{r,fig.height=8,fig.width=10 }
plots <- scGate::plot_levels(obj)
plots[["annotation"]] <- DimPlot(obj, group.by = "cell_type",label = T, repel =T,label.size = 3) + ggtitle("Author's manual annot") + NoLegend() 
wrap_plots(plots, ncol = 2)
```

## Evaluate performance of filtering compared to original annotation
```{r,collapse=T}
perf.scGate = scGate::performance.metrics(actual = obj$Plasma_cell, pred = obj$is.pure== "Pure")
perf.scGate
```





